---
title: "Tutorial 1" 
format: html 
editor: visual 
execute-dir: project
knitr:
  opts_chunk:
    engine.path: "C:/Program Files/Git/usr/bin/bash.exe"
---

## Assignment 1

You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.

First I am creating a directory to extract the assignment data into
```{bash}
cd /c/Users/zacha/gen_data

pwd # gen_data

# mkdir ./02_activities/data/assignment_1_data

```


Fixing an R chunk-specific working directory pathing issue
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```
Now I am going to load all necessary R libraries

```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(ggplot2)
library(dplyr)
library(seqminer)
library(HardyWeinberg)

```

(i) Read the .fam file. How many samples does the dataset contain?
```{bash, results='hide',message=FALSE, warning=FALSE}
cd /c/Users/zacha/gen_data
wc -l ./02_activities/data/assignment_1_data/gwa.qc.A1.fam # 4000

```         

4000 samples

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?
```{bash, results='hide',message=FALSE, warning=FALSE}
cd /c/Users/zacha/gen_data

head ./02_activities/data/assignment_1_data/gwa.qc.A1.fam

```         

Continuous

(iii) Read the .bim file. How many SNPs does the dataset contain?
```{bash, results='hide',message=FALSE, warning=FALSE}
cd /c/Users/zacha/gen_data
head ./02_activities/data/assignment_1_data/gwa.qc.A1.bim

wc -l ./02_activities/data/assignment_1_data/gwa.qc.A1.bim # 101083

```         

101083 SNPs

#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

Creating the genotype matrix
```{bash, results='hide',message=FALSE, warning=FALSE}
cd /c/Users/zacha/gen_data

"/c/Program Files (x86)/PLINK/plink2" --bfile ./02_activities/data/assignment_1_data/gwa.qc.A1 \
      --extract ./02_activities/data/assignment_1_data/snplist.txt \
      --export A \
      --out ./02_activities/data/assignment_1_data/geno.additive
```

Examining matrix in R to confirm --bfile worked as expected
```{r, message=FALSE, warning=FALSE}
# Load genotype data

geno <- fread("./02_activities/data/assignment_1_data/geno.additive.raw")
geno[1:5, 1:10]

```

Calculating allele frequencies
```{bash, results='hide',message=FALSE, warning=FALSE}
cd /c/Users/zacha/gen_data
"/c/Program Files (x86)/PLINK/plink2" --bfile ./02_activities/data/assignment_1_data/gwa.qc.A1 --freq \
      --extract ./02_activities/data/assignment_1_data/snplist.txt \
      --out ./02_activities/data/assignment_1_data/geno_freq
```

Examine allele freqencies
```{r, message=FALSE, warning=FALSE}
freq <- fread("./02_activities/data/assignment_1_data/geno_freq.afreq")
head(freq)

# Calculating allele frequencies

freq <- freq %>%
    mutate(ref_freq = 1-ALT_FREQS)

freq
```

rs3813199_G = 0.9430874
rs11804831_T = 0.8456590
rs3128342_C = 0.6948790
rs1861_C = 0.9460141

(ii) What are the minor allele frequencies (MAFs) for these four SNPs? 

The code for this was completed in the previous chunks

rs3813199_A = 0.0569126
rs11804831_C = 0.1543410
rs3128342_A = 0.3051210
rs1861_A = 0.0539859

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

Running the HWE test
```{bash, results='hide',message=FALSE, warning=FALSE}  
cd /c/Users/zacha/gen_data
"/c/Program Files (x86)/PLINK/plink2" --bfile ./02_activities/data/assignment_1_data/gwa.qc.A1 \
      --hardy \
      --out ./02_activities/data/assignment_1_data/gwa_qc_hwe
```

Now displaying the first few rows in R

```{r, message=FALSE, warning=FALSE}
hwe <- fread("./02_activities/data/assignment_1_data/gwa_qc_hwe.hardy")
head(hwe)

```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

```{r, message=FALSE, warning=FALSE}
snp_list <- c("rs1861", "rs3813199", "rs3128342", "rs11804831")

hwe[hwe$ID %in% snp_list, c(2, 10)]

```

rs1861, p = 0.274719
rs3813199, p = 1.000000
rs3128342, p = 0.330273
rs11804831, p = 0.113354

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```{r, message=FALSE, warning=FALSE}      
# Examine the SNP. The major allele is C
table(geno$rs1861_C)
freq[freq$ID == 'rs1861']

# Fit the linear regression model

model <- glm(PHENOTYPE ~ rs1861_C, data = geno)
summary(model)
model

```

The p value is <2.2e-16

(ii) How would you interpret the beta coefficient from this regression?

```{r, message=FALSE, warning=FALSE}
coefficients(model)

```

beta = 0.97382409

Phenotype increases with more copies of the rs1861_C allele 

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

```{r, message=FALSE, warning=FALSE}
p1 <- ggplot(geno, aes(x = rs1861_C, y = PHENOTYPE)) +
  geom_point(width = 0.1, alpha = 0.2, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1.2) +
  labs(title = "Linear Regression: PHENOTYPE ~ Genotype (Additive)",
       x = "Genotype (0/1/2)", y = "Phenotype") +
  theme_minimal()

print(p1)

```

(iv) Convert the genotype coding for rs1861 to recessive coding.

```{r, message=FALSE, warning=FALSE}
geno_rec <- geno  # make a copy
geno_rec[, 7:ncol(geno_rec)] <- as.data.frame(
  lapply(geno[, 7:ncol(geno)], function(x) ifelse(x == 2, 1, 0))
)
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```{r, message=FALSE, warning=FALSE}
table(geno_rec$rs1861_C)

# Fit the linear regression model

model_rec <- glm(PHENOTYPE ~ rs1861_C, data = geno_rec)
summary(model_rec)
model_rec         
```

p = <2.2e-16

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

```{r, message=FALSE, warning=FALSE}

p2 <- ggplot(geno_rec, aes(x = rs1861_C, y = PHENOTYPE)) +
  geom_point(width = 0.05, alpha = 0.2, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1.2) +
  labs(title = "Linear Regression: PHENOTYPE ~ Genotype (Recessive)",
       x = "Genotype (0/1)", y = "Phenotype") +
  theme_minimal()

print(p2)

```

(vii) Which model fits better? Justify your answer.

The additive model fits better as it has a lower AIC (11277) compared to the recessive model (11292). However, the magnitude of the difference in AIC is quite small, indicating that the additive model fits only marginally better.

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
